Report on the Third Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE3)

### 2015-12-23 WSSSPE3/report/WSSSPE3_review.tex
adding some thoughts to the abstract and conclusions - comments/edits welcome

\begin{abstract}
This technical report records and discusses the Third Workshop on Sustainable
Software for Science: Practice and Experiences (WSSSPE3). 
%The workshop used an
%alternative submission and peer-review process, which led to a set of papers
%divided across five topic areas: 
The report includes a description of the keynote presentation and a set of lightning
talks, and focuses on the discussions, future steps, and future organization for
a set of self-organized working groups.
For each group, the report also includes a point of contact and a landing
page that can be used by those who want to join that group's future activities.
\todo{think about if there are any general conclusions we want to highlight from WSSSPE3}
%%
%The workshop recognized that reliance on scientific software
%is pervasive in all areas of world-leading research today. The workshop
%participants then proceeded to explore different perspectives on the concept of
%sustainability. Key enablers and barriers of sustainable scientific software
%were identified from their experiences. In addition,
%recommendations with new requirements such as software credit files and software
%prize frameworks were outlined for improving practices in sustainable software
%engineering.
%%
%There was also broad consensus that formal
%training in software development or engineering was rare among the
%practitioners. Significant strides need to be made in building a sense of
%community via training in software and technical practices, on increasing their
%size and scope, and on better integrating them directly into graduate education
%programs.
%%
%Finally, journals can define and publish policies to improve reproducibility, whereas
%reviewers can insist that authors provide sufficient information and access to
%data and software to allow them reproduce the results in the paper. Hence a list of
%criteria is compiled for  journals to provide to reviewers so as to make it easier to
%review software submitted for publication as a ``Software Paper.''

@@

The working groups met for a small part of the first half day and all of the
second day, excepting some times for the groups to report back to the collected
workshop attendees. \todo{need to say some more about what came of this}

\todo{this paragraph is from the wssspe2 report - to be updated...} However, the
challenge that we have discovered since WSSSPE2 is that it is very hard to
continue the breakout groups' activities. The WSSSPE2 participants were willing
to dedicate their time to the groups while they were at the meeting, but
afterwards, they went back to their (paid) jobs. We need to determine how
to tie the WSSSPE breakout activities to people's jobs, so that they feel that
continuing them is a higher priority than it is now, perhaps through funding the
participants, or through funding coordinators for each activity, or perhaps by
getting the workshop participants to agree to a specific schedule of activities
during the workshop.

>>>>

\begin{abstract}
This technical report records and discusses the Third Workshop on Sustainable
Software for Science: Practice and Experiences (WSSSPE3). 
%The workshop used an
%alternative submission and peer-review process, which led to a set of papers
%divided across five topic areas: 
The report includes a description of the keynote presentation and a set of lightning
talks, and focuses on the discussions, future steps, and future organization for
a set of self-organized working groups.
For each group, the report also includes a point of contact and a landing
page that can be used by those who want to join that group's future activities.
The main challenge left by the workshop is to see if the groups will carry
out these activities that they have scheduled, and how the WSSSPE community
can encourage its members to do so.
%%
%The workshop recognized that reliance on scientific software
%is pervasive in all areas of world-leading research today. The workshop
%participants then proceeded to explore different perspectives on the concept of
%sustainability. Key enablers and barriers of sustainable scientific software
%were identified from their experiences. In addition,
%recommendations with new requirements such as software credit files and software
%prize frameworks were outlined for improving practices in sustainable software
%engineering.
%%
%There was also broad consensus that formal
%training in software development or engineering was rare among the
%practitioners. Significant strides need to be made in building a sense of
%community via training in software and technical practices, on increasing their
%size and scope, and on better integrating them directly into graduate education
%programs.
%%
%Finally, journals can define and publish policies to improve reproducibility, whereas
%reviewers can insist that authors provide sufficient information and access to
%data and software to allow them reproduce the results in the paper. Hence a list of
%criteria is compiled for  journals to provide to reviewers so as to make it easier to
%review software submitted for publication as a ``Software Paper.''

@@

The working groups met for a small part of the first half day and all of the
second day, with the exception of some short periods for the groups to report
back to the collected workshop attendees. Each group determined a set
of activities that the members could do to advance sustainable software in
a particular area.

The results of these group sessions made it clear that
there are many interlinked challenges in sustainable software, and that
while these challenges can be addressed, doing so is difficult because they
generally are not the full-time job of any of the attendees. 
As was the case in WSSSPE2 as well, the participants were willing
to dedicate their time to the groups while they were at the meeting, but
afterwards, they went back to their (paid) jobs.

We need to determine how
to tie the WSSSPE breakout activities to people's jobs, so that they feel that
continuing them is a higher priority than it is now, perhaps through funding the
participants, or through funding coordinators for each activity, or perhaps by
getting the workshop participants to agree to a specific schedule of activities
during the workshop as we have tried to do in WSSSPE3.  It remains to be
seen, however, if the participants will meet the schedules they set.

The overall challenge
left to the sustainable software community is perhaps one of organization: how
to combine the small partial efforts of a large number of people to impact a much
larger number of people: those who develop and use scientific software.  While
WSSSPE might help focus the actions of the groups, something more is needed
to incentivize the wider community, which is a generalization of the sustainable
software problem itself.
